---
title: 'STAT167 HW5 - Spring 2021'
author: 
- name: Deisy Diaz
  email: ddiaz022@ucr.edu
output:
  html_document:
    toc: yes
    toc_depth: 4
---

## **Homework #5 instructions**

**Review [the ISLR textbook](https://www.statlearning.com) Chapter 3 "Linear Regression", as well as lecture notes on map visualization (2021/5/4) and linear regression (2021/5/4 and 2021/5/6) before answering the homework questions**.

This homework contains 2 questions, each with multiple parts, 100 points in total.  

<font color="blue">Replace **INSERT_YOUR_ANSWER** with your own answers.</font>

- First open this `rmd` file in RStudio and click `Knit -> Knit to HTML` to render it to HTML format. 
- It is best to read this `rmd` file and the rendered `html` file side-by-side, while you are working on this homework.
- If the question asks you to write some R code, remember to put your code into a **R code chunk**. Make sure both your R code chunk and the output messages / figures are visible in the rendered `html` file.
- For this homework, use **ggplot2** to visualize your data.

- <font color="blue">**Please comment your R code thoroughly, and follow the R coding style guideline (https://google.github.io/styleguide/Rguide.xml). Partial credit will be deducted for insufficient commenting or poor coding styles.**</font>

- If you have any question about this homework assignment, we encourage you to post it on [**Piazza**](http://piazza.com/ucr/spring2021/stat167/home).

### **Homework submission guideline**

- **This homework is DUE at _11:59 PM_ on _Sunday May 9th, 2021_**.
- Late submission penalties.  
    - Submissions up to 24 hours late will incur a 10% deduction.  
    - Submissions up to 48 hours late will incur a 30% deduction.

- <font color="blue">**If you are using one or both of your free late days, please state here: **</font> **INSERT_YOUR_ANSWER**

- After you complete all questions, save your `rmd` file to `FirstnameLastname-SID-HW5.rmd` and save the rendered `html` or `pdf` file to `FirstnameLastname-SID-HW5.html` or `FirstnameLastname-SID-HW5.pdf`.
- Submit **BOTH your source `rmd` file and the knitted `html` or `pdf` file** to **GradeScope**. Do NOT create a zip file. You can submit multiple times, you last submission will be graded.

***

## **Acknowledgements**

<font color="blue">Please list all the help you have received for completing this homework.</font>

**INSERT_YOUR_ANSWER**

***

**Install necessary packages**

Note that you only need to install each package once. Then you can comment out the following installation lines.

``````{r}
#install.packages("maps")
#install.packages("mapproj")
```

***

**Load necessary packages**

```{r, collapse=T}
library(tidyverse) # for `ggplot2`, `dplyr`, and more

library(maps) # for map visualization
#library(mapdata)
library(mapproj) 

library(datasets) # for `state` data set
library(nycflights13) # for the 2013 NYC flights data set
```

***

## **Question 1 [45pt] Map visualization with `ggplot2`**

Recall in HW1, we used the `maps` package to generate map visualizations. In this homework, we will draw maps with `ggplot2`. 

The `maps` package comes with a plotting function, but, we will opt to use `ggplot2` functions (`geom_polygon` and `geom_map`) to plot the maps in the `maps` package.

Recall that `ggplot2` operates on data frames. We will use the `map_data()` function (provided by `ggplot2`), which turns a series of points along an outline into a data frame of those points.

### (a) [5pt] Plot the USA map

First we load the USA map from `maps`.
```{r, collapse=T}
usa_map <- map_data("usa")
dim(usa_map)
glimpse(usa_map)
```

In the lecture, we have learned how to use `geom_polygon()` to make a simple black map (no line color, but with a black fill).

```{r}
ggplot(data = usa_map) + 
  geom_polygon(aes(x = long, y = lat, group = group)) + 
  coord_quickmap()
```

Google the coordinates of UCR, and then use `geom_point()` to mark the location of UCR on the USA map.
In addition, change the outline/border color as well as the fill-in color of your map.

**ANSWER**

```{r}
# plot data by usa_map data by long, lat, group by group, fill color green and color brown then plot point of ucr coordinates
ggplot(data = usa_map) + 
  geom_polygon(aes(x = long, y = lat, group = group), fill = "green", color = "brown") + 
  coord_quickmap() + 
  geom_point( mapping = aes(x = -117.33, y = 33.97), color = "red")

```

***

### (b) [15pt] Plot the states map

In addition to `geom_polygon()`, we can use `geom_map()` to draw maps too. Basically, `geom_map()` acts as a wrapper of `geom_polygon()`. See more details in the `geom_map()` documentation @ http://ggplot2.tidyverse.org/reference/geom_map.html

Here is the example code of a states map from the lecture. 
We can plot all the states, map the `fill` aesthetic to `region` and set the the lines of state borders to white color.

```{r, collapse=T}
states_map <- map_data("state")
dim(states_map)
glimpse(states_map)
```

```{r, collapse=T}
ggplot() + 
  geom_map(data = states_map, map = states_map,
           aes(map_id = region, fill = region), color="white") + 
  # geom_map() doesn't work in such a way that ggplot2 knows the extend of the map values, so you always have to expand the limits by hand yourself.
  expand_limits(x = states_map$long, y = states_map$lat) +
  coord_quickmap() +
  guides(fill = FALSE) # do this to leave off the color legend
```

Next, we will use the built-in `state` data sets in R to annotate our states map. In particular, `state.x77` is a two-dimensional array containing 8 variables and data from all 50 states.

```{r, collapse=T}
#?state.x77
head(state.x77)

state_data <- as.data.frame(state.x77)
state_data$State <- tolower(rownames(state_data))
state_data %>% glimpse()
```

Below is the example code from the lecture for a state population map. We first create an aesthetic mapping for `map_id` to the column `State` (state names in lower case) in the `state_data` data frame. We then call `geom_map` again and map the `fill` aesthetic to the `Population` variable in `state_data`.

```{r, collapse=T}
ggplot(data = state_data, aes(map_id = State))+ 
  geom_map(map = states_map,
           aes(fill = Population)) + 
  expand_limits(x = states_map$long, y = states_map$lat) +
  coord_quickmap() 
```

#### (i) [5pt] Draw your own states map and map the `fill` aesthetic to `Income` in the `state.x77` data set

**ANSWER**

```{r}
# plot map of us states and fill by income
ggplot(data = state_data, aes(map_id = State))+ 
  geom_map(map = states_map,
           aes(fill = Income)) + 
  expand_limits(x = states_map$long, y = states_map$lat) +
  coord_quickmap() 

```

***

#### (ii) [10pt] Add 50 colorful points to your map 
Use one point to mark one state (state coordinates can be found in `state.center`). Map the color of the points to `state.region`. Map the `size` aesthetic of the points to `Population`.

```{r, collapse=T}
head(state.center)
head(state.region)
```

**INSERT_YOUR_ANSWER**

```{r}
# turn it into a data frame
state_centers <- as.data.frame(state.center)
# plot data with x = x and y = y color by region and size by income
ggplot(data = state_data, aes(map_id = State))+ 
  geom_map(map = states_map,
           aes(fill = Income)) + 
  expand_limits(x = states_map$long, y = states_map$lat) +
  geom_point(aes(x = state_centers$x, y = state_centers$y, color = state.region, size = state_data$Income)) +
  coord_quickmap() 

```

***

### (c) [25pt] NYC flight destination map

Recall `nycflights13::flights` data set contains all 336,776 flights that departed from New York City in 2013.

```{r, collapse=T}
#?flights # full documentation
glimpse(flights)
```

#### (i) [10pt] Count the number of flights per destination

How many unique destination airports did these NYC flights connected to? How many **non-canceled** flights per destination? Which destination had the largest number of arrival flights from NYC? Which destination had the smallest number of arrival flights from NYC?

**Note:** Recall in HW4 we have cleaned the `flights` data by removing flight record that has missing values in `dep_delay` or `arr_delay`, and save the result as the **non-canceled** flights.

**INSERT_YOUR_ANSWER**

```{r}
all_flights_count <- flights %>%
  group_by(carrier, dest) %>%
  count(carrier, dest)
all_flights_count
# There are a total of 314 unique destinations

not_canceled <- filter(flights, !is.na(dep_delay), !is.na(arr_delay))
noncanceled_flight_count <- 
  not_canceled %>%
  group_by(carrier, dest) %>%
  count(carrier, dest)
# Display the number of flights per destination

noncanceled_flight_count %>%
  arrange(desc(n))
# The largest number of flights were to ATL

noncanceled_flight_count %>%
  arrange(n)
# arrange flights by n from least to greatest
```


*** 

#### (ii) [15pt] Mark all destination airports on a states map
Find out the coordinates of the destination airports from `nycflights13::airports`. Draw each destination airport as a point on a states map, and map a point aesthetic to the number of **non-canceled** flights flew to that destination from NYC in 2013.

**Hint:** Suppose that you have saved your list of unique destinations and their corresponding flight counts in a tibble called `per_dest`. Then you can _join_ `per_dest` and `airports` using destination airport FAA code as the _key_.

```{r}
#?airports
per_dest <- # create per_dest by using not_canceled data group by carrier and dest then count
  not_canceled %>% 
  group_by(carrier, dest) %>%
  count(carrier, dest)

airports %>% glimpse()
all_data <- left_join(per_dest, airports, by=c("dest"="faa")) # create a data frame from per_dest and airports
all_data # print to check
```

**ANSWER**

```{r}

# Graph polygon of states map with states white border, map points from all_data x = long y = lat size = count of each non canceled flight
ggplot() + 
  geom_polygon(data = states_map, aes(x = long, y = lat, fill = region), color = "white") + 
  expand_limits(x = states_map$long, y = states_map$lat) +
  geom_point(aes(x = all_data$lon, y = all_data$lat, size = per_dest$n)) +
  coord_quickmap() +
  guides(fill = FALSE)
  

```

Pay attention to the warning message. Which destination airports have missing values?

**INSERT_YOUR_ANSWER**
```{r}
# grop data by dest and then find missing values
all_data %>%
  group_by(dest) %>%
  filter(is.na(lon))

# There are four destinations with NA's. SJU, STT, BQN, and PSE.

```

***

## **Question 2 [55pt] Simple linear regression**

### (a) [10pt] Population regression line vs least squares regression line

In this question, we will reproduce the left panel of Figure 3.3 in the **ISLR** textbook.

Write your own R code the simulate 100 data points from the following model.

$$Y=2+3X+\epsilon$$
$$X \sim Uniform(A=-2, B=2)$$
$$\epsilon \sim N(\mu=0, \sigma=2)$$

Then use `ggplot2` to make a scatter plot of the 100 data points $(x_1,y_1), \cdots (x_{100},y_{100})$. Add the population regression line $Y=2+3X$ in red color; add the least squares fit of your simulated data as a blue line. 

**INSERT_YOUR_ANSWER**

```{r, collapse=T}

# Always remember to set the random seed for simulation studies
set.seed(167) # feel free to change 167 to your lucky number
# generate x
x <- runif(100, -2, 2)
# generate e
e <- rnorm(100, 0, 2)
# calculate y
y <- 2 + 3*x + e
# make a dataframe of x and y
data1 <- as.data.frame(cbind(x,y))
# graph population regression line in red and least squares fit in blue
ggplot(data1, aes(x = x, y = y)) +
  geom_point() +
  geom_abline(slope = 3, intercept = 2, color = "red") + 
  geom_smooth(method = "lm", se = F, color = "blue") +
  coord_cartesian(xlim = c(-2,2), y = c(-10,10))

```

***

### (b) [10pt] Estimate the coefficients $\beta_0$ and $\beta_1$ 

In the lecture, we have learned that the analytical solution of the least squares regression $Y = \beta_0 + \beta_1 X + \epsilon$ is

$$ 
\begin{aligned}
 \hat{\beta_1} & = 
 \frac{\sum_{i=1}^{n} (x_i - \bar{x})(y_i - \bar{y})}
 {\sum_{i=1}^{n} (x_i - \bar{x})^2} \\
 \hat{\beta_0} & = \bar{y} - \hat{\beta_1} \bar{x}
\end{aligned} 
$$

Write your own R code to calculate $\hat{\beta_0}$ and $\hat{\beta_1}$ using the above equations. Compare your results with the coefficients calculated by the `lm` function. Are they the same?

**ANSWER**
```{r}
# calculate beta1
b1 <- sum((x-mean(x))*(y-mean(y)))/(sum((x-mean(x))^2))
#calculate beta0
b0 <- mean(y)- b1*mean(x)
b1 # print
b0 # print

regress <- lm(y~x, data1) # save linear model of y to x 
summary(regress) # print the summary

# Yes, they are the same
```

***

### (c) [20pt] Assess accuracy of the regression model

Recall that the accuracy of a linear regression fit is typically assessed using the **residual standard error (RSE)** and the **$R^2$ statistic**.

#### (i) [10pt] Calculate the residual standard error (RSE)

Residual standard error (RSE) is an estimate of $\sigma$, the standard deviation of $\epsilon$.

$$
\begin{aligned}
\text{RSE} = \hat{\sigma} &= \sqrt{\frac{1}{n-2}\text{RSS}} \\
&= \sqrt{\frac{1}{n-2} \sum_{i=1}^{n} ( y_i - \hat{y_i} )^2} 
\end{aligned}
$$ 
Write your own R code to calculate RSE using the above equation. Compare your result with the residual standard error calculated by the `lm` function. Are they the same? How does the RSE value relate to our noise model $\epsilon \sim N(\mu=0, \sigma=2)$?

**INSERT_YOUR_ANSWER**

```{r}
# calculate rss
rss <- sum((y-b0-b1*x)^2)
# calculate rse
rse <- sqrt((1/98)*rss)
rse # print rse
summary(regress) # summary of regress

# yes they are the same.
```


***

#### (ii) [10pt] Calculate the $R^2$ statistics

Write your own R code to compute the $R^2$ value using the following equations.

$$
\begin{aligned}
\text{TSS} & = \sum_{i=1}^{n} (y_i - \bar{y})^2 \\
\text{RSS} & = \sum_{i=1}^{n} \left(y_i - \hat{f}(x_i) \right)^2 \\
R^2 & = \frac{\text{TSS}-\text{RSS}}{\text{TSS}} = 1 - \frac{\text{RSS}}{\text{TSS}} \\
\end{aligned}
$$
Compare your result with the R-squared value calculated by the `lm` function. Are they the same? 

**INSERT_YOUR_ANSWER**

```{r}

tss <- sum((y - mean(y))^2) # calcuate tss
rsquared <- (tss-rss)/tss # calculate rsquared
rsquared # print r squared
summary(regress) # print summary of linear model

# They are the same
```

***

### (d) [15pt] Simulations of least squares regression

Now, let's repeat the simulation 10 times, and reproduce the right panel of Figure 3.3 in the **ISLR** textbook.

**Notes**: 

- You can draw all the least squares lines using the same blue color.
- Write your code using a `for` loop, rather than manually copying and pasting your code 10 times.
- You can save your ggplot to a variable (e.g. `gg`) and keep adding new layers inside your for loop. So the pseudocode will look like:

```{r}
# plot population regression line
gg <- ggplot() +
  geom_abline(slope = 3, intercept = 2, color = "red") 
  #n <- 1

for ( i in 1:10 ) { # run for lop 10 times
  #n <-  n+1
  x <- runif(100, -2, 2) # calculate x
  e <- rnorm(100, 0, 2) # calculate e

  y <- 2 + 3*x + e # calculate y
  data2 <- as.data.frame(cbind(x,y)) # save data into a dataframe
  gg <- gg +  # plot data using linear model
    geom_smooth(data = data2, aes(x, y), method = "lm", se = F) ## alternatively you can use geom_abline()
}

gg ## call gg at the end to plot it.
```

***