---
title: 'STAT167 Lab #1 - Spring 2021'
author: 
- name: Deisy Diaz Vazquez
  email: ddiaz022@ucr.edu
date: "2021/4/2"
output:
  html_document:
    toc: yes
---

**Acknowledgment**: Part of this R Markdown template is adapted from David Dalpiaz (UIUC).

# **Discussion week 1 instructions**
This week, you will first learn the basic R Markdown syntax. 

- First, download the `rmd` file from iLearn. It gives you an option to "open" or "save". Make sure that you **save** the file to the Desktop, Downloads, Documents, or any other location where you have access to.
- Open this `rmd` file in RStudio and click `Knit -> Knit to HTML` to render it to HTML format. (If you have `LaTex` installed on your computer you can also render it to PDF format).
- Read this `rmd` file and the rendered `html` file side-by-side, to see how this document was generated!
- Be sure to play with this document! Change it. Break it. Fix it. The best way to learn R Markdown (or really almost anything) is to try, fail, then find out what you did wrong.

Next, you will review some example code from this week's lectures. 

- Read over the code and the output. If you have any questions about certain functions or parameters, it is the time to ask!
- There are some exercises through out this document. Replace **INSERT_YOUR_ANSWER** with your own answers. Knit the file to HTML, and check your results.

<font color="blue">**Please comment your R code thoroughly, and follow the R coding style guideline (https://google.github.io/styleguide/Rguide.xml). Partial credit will be deducted for insufficient commenting or poor coding styles.**</font>

**Lab submission guideline**

- After you completed all exercises, save your file to `FirstnameLastname-SID-lab1.rmd` and save the rendered html/pdf file to `FirstnameLastname-SID-lab1.html` or `FirstnameLastname-SID-lab1.pdf`. 
- Submit **BOTH your source `rmd` file and the knitted `html` or `pdf` file** to **GradeScope**. Do NOT create a zip file. You can submit multiple times, you last submission will be graded.


# RMarkdown Basics

[RMarkdown](http://rmarkdown.rstudio.com/) at its core is a combination of [`R`](https://cran.r-project.org/) and [Markdown](https://daringfireball.net/projects/markdown/) used to generate reproducible reports for data analyses.

Markdown and `R` are mixed together in a `.rmd` file, which can then be rendered into a number of formats including `.html`, `.pdf`, and `.docx`. There is a strong preference for using `.html` in this course. 

Have a look at this `.rmd` to see how this document was generated! It should be read alongside the rendered `.html` to best understand how everything works. You can also modifying the `.rmd` along the way, and see what effects your modifications have.

Formatting text is easy. **This is bold.** *This is italics.* `This text appears as monospaced.`

# Test Header 1
## Test Header 2
### Test Header 3

- Unordered list element 1.
- Unordered list element 2.
- Unordered list element 3.

1. Ordered list element 1.
2. Ordered list element 2.
3. Ordered list element 3.

# Packages

Packages are key to using `R`. The community generated packages are a large part of `R`'s success, and it is extremely rare to perform an analysis without using at least some packages. Once installed, packages must be loaded before they are used.

## Install necessary packages

```{r, echo=T}
# If you can knit the html file successfully, do not change this R code chunk

# Otherwise, uncomment the two lines and install the packages.
# You only need to run the install.packages() commands once. 
# Then you can comment them out by adding `#` back at the beginning of each line

#install.packages("rmarkdown", repos="http://cran.rstudio.com/")
#install.packages("yaml", repos="http://cran.rstudio.com/")
```

Note that `rmarkdown` is actually a package in `R`! If `R` never prompts you to install `rmarkdown` and its associated packages when first creating an RMarkdown document, use the above command to install them manually.

# Adding `R`

So far we have only used Markdown to create html. This is useful by itself, but the real power of RMarkdown comes when we add `R`. There are two ways we can do this. We can use `R` code chunks, or run `R` inline.

## `R` Code Chunks

The following is an example of an `R` code chunk

```{r example_chunk}
# generate random normals
set.seed(167) # feel free to change 167 to your lucky number
x <- rnorm(100) # use <- for assignment instead of =
```

There is a lot going on here. In the `.rmd` file, notice the syntax that creates and ends the chunk. Also note that `example_chunk` is the chunk name. Everything between the start and end syntax must be valid `R` code. Chunk names are not necessary, but can become useful as your documents grow in size.

## Inline `R`

`R` can also be ran in the middle of exposition. For example, the mean of the data we generated is `r mean(x)`.

## Chunk Options

There are many [chunk options](http://yihui.name/knitr/options/). Here we first introduce two options which are frequently used: `eval` and `echo`.

```{r not_run, eval = FALSE}
?log
x
```

Using `eval = FALSE` the above chunk displays the code, but it is not run. The `?` code pulls up documentation of a function. The `x` code prints all values inside the vector x. With `eval = FALSE` (or simply `eval = F`), there is no output displayed.

```{r not_displayed, echo = FALSE}
print("Hello World!")
```

Above, we see output, but no code! This is done using `echo = FALSE` (or simply, `echo=F`), which is often useful.

# Adding Math with `LaTeX`

Another benefit of RMarkdown is the ability to add [Latex for mathematics typesetting](https://www.latex-project.org/about/). Like `R` code, there are two ways we can include Latex; displaystyle and inline.

Note that use of `LaTeX` is somewhat dependent on the resulting file format. For example, it cannot be used at all with `.docx`. To use it with `.pdf` you must have LaTeX installed on your machine.

With `.html` the `LaTeX` is not actually rendered during knitting, but actually rendered in your browser using MathJax.

## Displaystyle `LaTeX`

Displaystyle is used for larger equations which appear centered on their own line.

$$
s = \sqrt{\frac{1}{n - 1}\sum_{i=1}^{n}(x_i - \bar{x})^2}
$$

## Inline LaTex

We could mix LaTeX commands in the middle of exposition, for example: $t = 2$. We could actually mix `R` with Latex as well! For example: $\bar{x} = `r mean(x)`$.

# **Lecture Review**

## Generate sample data

Recall that in the lecture, we generated 100 random normals, assigned the values into a vector `x`, then printed out some values in vector `x`.

```{r}
# generate random normals
set.seed(167) # feel free to change 167 to your lucky number
x <- rnorm(100) # use <- for assignment instead of =
# print out the data or part of it
x
1:10
10:1
seq(from=1, to=10, by=2)
seq(10, 1)
x[1:10]
y <- c(2,4,7)
y
x[y]
length(x)
length(x[y])
```

### **Exercise #1** 

Look at the output of the following code, what happens when you select with negative index?

```{r}
x[4]
x[-4]
```

**ANSWERS**: The command *x[-4]* will display all the values in the vector except the last four.

## Calculate summary statistics

We can calculate some summary statistics from the data. Here we set the option `collapse=T` for our code chunk, so that R Markdown will try to collapse all the source and output blocks from one code chunk into a single block.

```{r, collapse=T}
mean(x)
median(x)

var(x) # sample variance
sd(x) # sample standard deviation

n <- length(x)
sum((x-mean(x))^2)/n # population variance
mean(x^2)-(mean(x))^2 # population variance
var(x)*(n-1)/n # population variance
sd(x)^2*(n-1)/n # population variance

summary(x)
```

### **Exercise #2** 

Complete the following code to print out the maximum value and the minimum value in `x`.

**ANSWERS**

```{r}
# INSERT_YOUR_ANSWER
min(x) # this will give the min value in x
max(x) # this will give the max value in x
```

### **Exercise #3** 

The `sd()` function calculate the sample standard deviation. 
$$ s = \sqrt{\frac{1}{n-1}\sum_{i=1}^{n}(x_i - \bar{x})^2}$$
How to calculate the population standard deviation?
$$ \sigma = \sqrt{\frac{1}{n}\sum_{i=1}^{n}(x_i - \bar{x})^2}$$

**ANSWERS**

```{r, collapse=T}
sd(x) # sample standard deviation s

# calculate the population standard deviation
# INSERT_YOUR_ANSWER
sqrt(sum((x-mean(x))^2)/n)
```

## Plot single data set

Note that you can modify chunk options `fig.height` and `fig.width` to change the size of output plots from a particular chunk.

```{r, fig.height=4, fig.width=10, fig.align = "center"}
par(mfrow=c(1,3)) # 1-by-3 layout
# subfigure #1
boxplot(x, horizontal=T, col=2, border="blue", xlab="Random Numbers", main="Boxplot")

# subfigure #2
hist(x, xlab="Random Numbers", main="Histogram")

# subfigure #3
plot(density(x), main="Density Curve")
```
