---
title: "STAT167 Lab #4 - Spring 2021"
author: 
- name: Deisy Diaz Vazquez
  email: ddiaz022@ucr.edu
date: "2021/4/23"
output:
  html_document:
    toc: yes
---

# **Discussion week 4 instructions**

This week, we will review some `ggplot2` example figures. 


- First, download the `rmd` file from iLearn. 
- Open this `rmd` file in RStudio and click `Knit -> Knit to HTML` to render it to HTML format. (The computer lab computers do not have `LaTex` installed so you can not render it to PDF format).
- Read this `rmd` file and the rendered `html` file side-by-side, to see how this document was generated!
- Be sure to play with this document! Change it. Break it. Fix it. The best way to learn R Markdown (or really almost anything) is to try, fail, then find out what you did wrong.

- Read over the `ggplot2` example code and check the output. If you have any questions about certain functions or parameters, it is the time to ask!
- There are some exercises through out this document. Replace **INSERT_YOUR_ANSWER** with your own answers. Knit the file to HTML, and check your results.

<font color="blue">**Please comment your R code thoroughly, and follow the R coding style guideline (https://google.github.io/styleguide/Rguide.xml). Partial credit will be deducted for insufficient commenting or poor coding styles.**</font>

**Lab submission guideline**

- After you completed all exercises, save your file to `FirstnameLastname-SID-lab4.rmd` and save the rendered html/pdf file to `FirstnameLastname-SID-lab4.html` or `FirstnameLastname-SID-lab4.pdf`.
- Submit **BOTH your source `rmd` file and the knitted `html` or `pdf` file** to **GradeScope**. Do NOT create a zip file. You can submit multiple times, you last submission will be graded.

***

# **Lecture Review - `ggplot2`**

## Install the The `tidyverse` package

```{r, collapse=T}
# install the tidyverse package first if you have not done it  yet.
#install.packages("tidyverse") # you can comment out this line after you have installed `tidyverse`
library(tidyverse)
```

## The `mpg` data set

This data set contains fuel economy data 1999 - 2008 for 38 popular car models (source: EPA. [fueleconomy.com/](http://fueleconomy.gov)).

```{r, collapse=T}
?mpg
dim(mpg)  # dimension of the table
mpg  # print/view mpg (we will introduce tibble later in this class)
str(mpg)  # list the structures in mpg
glimpse(mpg) # get a glimpse of the mpg data
```


## The complete graphing template in `ggplot2`
```{r, eval=F}
ggplot(data = <DATA>) + 
  <GEOM_FUNCTION>(
     mapping = aes(<MAPPINGS>),
     stat = <STAT>,        # optional
     position = <POSITION> # optional
  ) +
  <COORDINATE_FUNCTION> +  # optional
  <FACET_FUNCTION> +       # optional
  <SCALE_FUNCTION> +       # optional
  <THEME_FUNCTION>         # optional
```

## Example: `geom_boxplot()` for one variable

Let's first take a look at the distribution of highway mileage.

```{r, fig.height=5.5}
ggplot(data = mpg) +
  geom_boxplot(mapping = aes(x = "", y = hwy)) 
```

### **Exercise #1**

If you add `facet_wrap(~class)` to the above code, what will happen?

**ANSWER:** Adding that to the code above, will make a boxplot for each type of car class. Like so below.

```{r}

ggplot(data = mpg) +
  geom_boxplot(mapping = aes(x = "", y = hwy)) +
  facet_wrap(~class)

```

## Example: `geom_boxplot()` for two variables

How highway mileage varies across car classes?

- Plot the distribution of highway mileage by class.

```{r, fig.height=5.5}
ggplot(data = mpg) +
  geom_boxplot(mapping = aes(x = class, y = hwy))  
```

To make the trend easier to see, we can reorder `class` based on the median value of `hwy`.

```{r, fig.height=5}
ggplot(data = mpg) +
  geom_boxplot(mapping = aes(x = reorder(class, hwy, FUN = median), 
                             y = hwy))
```

When you have long variable names, `geom_boxplot()` will work better if you flip it with `coord_flip()`.

```{r, fig.height=5}
ggplot(data = mpg) +
  geom_boxplot(mapping = aes(x = reorder(class, hwy, FUN = median), 
                             y = hwy)) +
  xlab("class") +
  coord_flip()
```  

### **Exercise #2**

Can you draw a violin plot to compare the distribution of highway mileage varies across car classes? Order your violins by the median values of highway mileage and flip the plot to horizontal. 

**ANSWER**
```{r}

ggplot( data = mpg)+
  geom_violin(mapping = aes(x = reorder(class, hwy, FUN = median),
                            y = hwy))

```

### **Exercise #3**

There are two ways to zoom in a ggplot. Compare the following two plots. What are the differences? Which zooming option do you prefer? Explain your reasons.

**ANSWER**

```{r, collapse=T, fig.height=5}
ggplot(data = mpg) +
  geom_boxplot(mapping = aes(x = class, y = hwy)) +
  ylim(c(20, 30))
```  

```{r, collapse=T, fig.height=5}
ggplot(data = mpg) +
  geom_boxplot(mapping = aes(x = class, y = hwy)) +
  coord_cartesian(ylim=c(20, 30))
```  

## Example: `geom_histogram()` to compare highway mileage across car classes

```{r, collapse=T, fig.height=5}
ggplot(data = mpg) +
  geom_histogram(mapping = aes(x = hwy, fill = class),
                 binwidth = 10,
                 position = "dodge")
```

### **Exercise #4**

(a) Why there are gaps in the histogram?

**ANSWER** Because we use "dodge" but certain class of cars don't have mpgs in that branch.

(b) Can you change this frequency histogram to a density histogram?

**ANSWER**

```{r}

ggplot(data = mpg) +
  geom_histogram(mapping = aes(x = hwy, y = ..density.., fill = class),
                 binwidth = 10,
                 position = "dodge")

```

## Example: `geom_freqpoly()` to compare highway mileage across car classes

```{r, collapse=T, fig.height=5}
ggplot(data = mpg) +
  geom_freqpoly(mapping = aes(x = hwy, color = class),
                binwidth = 10)
```