---
title: "STAT167 Lab #5 - Spring 2021"
author: 
- name: "Deisy Diaz "
  email: "ddiaz022@ucr.edu"
date: "2021/4/30"
output:
  html_document:
    toc: yes
---

# **Discussion week 5 instructions**

This week, we will review how to transform/manipulate the data using `dplyr`.

- First, download the `rmd` file from iLearn. 
- Open this `rmd` file in RStudio and click `Knit -> Knit to HTML` to render it to HTML format. (If you have `LaTex` installed on your computer, you can also render it to PDF format).
- Read this `rmd` file and the rendered `html` file side-by-side, to see how this document was generated!
- Be sure to play with this document! Change it. Break it. Fix it. The best way to learn R Markdown (or really almost anything) is to try, fail, then find out what you did wrong.

- Read over the `dplyr` example code and the output. If you have any questions about certain functions or parameters, it is the time to ask!
- There are some exercises through out this document. Replace **INSERT_YOUR_ANSWER** with your own answers. Knit the file to HTML, and check your results.

<font color="blue">**Please comment your R code thoroughly, and follow the R coding style guideline (https://google.github.io/styleguide/Rguide.xml). Partial credit will be deducted for insufficient commenting or poor coding styles.**</font>

**Lab submission guideline**

- After you completed all exercises, save your file to `FirstnameLastname-SID-lab5.rmd` and save the rendered html/pdf file to `FirstnameLastname-SID-lab5.html` or `FirstnameLastname-SID-lab5.pdf`.
- Submit **BOTH your source `rmd` file and the knitted `html` or `pdf` file** to **GradeScope**. Do NOT create a zip file. You can submit multiple times, you last submission will be graded.

***

# **Lecture Review - data transformation with `dplyr`**

## Load the `tidyverse` package

```{r, collapse=T}
# If you don't have `tidyverse` on your computer, you need install it first. Then you can comment out the following line after the installation.
#install.packages("tidyverse") 

library(tidyverse)
```

## The `nycflights13::flights` data set

This data frame contains all 336,776 flights that departed from New York City in 2013. The data comes from the [US Bureau of Transportation Statistics](http://www.transtats.bts.gov/DatabaseInfo.asp?DB_ID=120&Link=0).

```{r}
# You need to install `nycflights13` first, then you can comment out the following line
#install.packages("nycflights13")
```

```{r, collapse=T}
library(nycflights13)
?flights # full documentation
# View(flights) # see the data in RStudio Viewer
flights
```

There are six commonly used data transformation functions in `dplyr`. The examples are reviewed below.

## Pick observations by their values - `filter()`

```{r, collapse=T}
# flights that departed on Jan/1/2013
filter(flights, month == 1, day == 1)
```

```{r, collapse=T}
# flights that departed in Nov or Dec
filter(flights, month == 11 | month == 12)

filter(flights, month %in% c(11, 12))
```

```{r, collapse=T}
# flights that weren't delayed (on arrival or departure) by more than two hours
filter(flights, !(arr_delay > 120 | dep_delay > 120))

filter(flights, arr_delay <= 120, dep_delay <= 120)
```

### **Exercise #1**

(a) Find the flights that departed in the summer (July, August, and September)

**ANSWER**

```{r}
# Filter flights by months July, August, and September
filter(flights, month %in% c(7,8,9))

```

(b) Find the flights that flew to Houston (IAH or HOU)

**ANSWER**

```{r}
# Filter flights of which destination is houston 
filter(flights, dest == "HOU" | dest == "IAH")

```

## Reorder the rows - `arrange()`

```{r, collapse=T}
# sort flights by date
arrange(flights, year, month, day)

# sort flight by departure delay in a descending order
arrange(flights, desc(dep_delay))
```

### **Exercise #2**

(a) Find the flights that traveled the shortest distance

**ANSWER**

```{r}
 # Arrange flights by distance in ascending order
arrange(flights, distance)

```

(b) Find the flights that had the shortest air time

**ANSWER**

```{r}
# Arrange flights by air time in ascending order
arrange(flights, air_time)

```

## Pick variables by their names - `select()`

```{r, collapse=T}
# select all columns between year and day (inclusive)
select(flights, year, month, day)

select(flights, year:day) 
```

```{r, collapse=T}
# select all columns except those from year to day (inclusive)
select(flights, -year, -month, -day)

select(flights, -(year:day)) 
```

There are a number of functions you can use within `select()`:

- `starts_with("abc")`: matches names that begin with "abc".

- `ends_with("xyz")`: matches names that end with "xyz".

- `contains("ijk")`: matches names that contain "ijk".

- `matches("(.)\\1")`: selects variables that match a regular expression. This one matches any variables that contain repeated characters.

- `num_range("x", 1:3)`: matches x1, x2 and x3. It is useful when numbers were included in column names.

- `one_of(...)`: selects columns names that are from a group of names. It is useful when columns are named as a vector or character string.

- `everything()`: selects all columns.

See `?select` for more details.

## Create new variables - `mutate()`

```{r, collapse=T}
# select a subset of variables to be used in the next step
flights_sml <- select(flights, month:day, ends_with("delay"), 
                      distance, air_time)
flights_sml

# create two new variables, gain and speed
mutate(flights_sml, 
       gain = arr_delay - dep_delay,
       speed = distance / air_time * 60)

# If you only want to keep the new variables, use `transmute()`
transmute(flights, 
          gain = arr_delay - dep_delay,
          hours = air_time / 60,
          gain_per_hour = gain / hours)
```

### **Exercise #3**

Find the fastest flight (i.e., the flight has the highest speed). 
Report the flight date, actual departure and arrival time, origin, destination, carrier, and flight number. 

**ANSWER**

```{r}
# create a new column in flights to calcukate speed
flights1 <- mutate(flights, 
                   speed = distance / air_time * 60)
arrange(flights1, desc(speed)) # arrange new data frame by speed
# Answer: 
# Flight date: 5/25/2013
# Departure: 5:09; Arrival: 7:23
# Origin: LGA; Destination: ATL
# Carrier: DL
# Flight number: 1499
```

## Grouped summaries with `group_by()` and `summarise()`

```{r, collapse=T} 
# `summarise()` collapses a data frame to a single row:
summarise(flights, delay = mean(dep_delay, na.rm = TRUE))

# `summarize()` is useful when pair it with `group_by()`
by_day <- group_by(flights, year, month, day)
summarise(by_day, delay = mean(dep_delay, na.rm = TRUE))
```

## Combining multiple operations with the pipe `%>%`

Imagine that we want to explore the relationship between the distance and average delay for each destination.

**Data transformation &  cleaning - step by step**

1. Filter out canceled flights;
2. Group non-canceled flights by destination;
3. Summarise grouped data to compute the number of flights, average distance, and average arrival delay per destination;
4. Filter out outliers (longest distances - HNL and ANC);
5. Filter out noisy points (small flight counts - rare destinations).

```{r}
# original code - step by step
not_canceled <- filter(flights, !is.na(dep_delay), !is.na(arr_delay))
not_canceled_by_dest <-  group_by(not_canceled, dest)
delays_by_dest <- summarise(not_canceled_by_dest, count = n(), 
                            dist = mean(distance), delay = mean(arr_delay))
delays_by_dest_filtered <- filter(delays_by_dest, 
                                  dest != "HNL", dest != "ANC")
delays_by_dest_cleaned <- filter(delays_by_dest_filtered, count >= 10)
```

```{r}
# alternative code using piping %>%
not_canceled <- filter(flights, !is.na(dep_delay), !is.na(arr_delay))
delays_by_dest_cleaned <- not_canceled %>% 
  group_by(dest) %>% 
  summarise(count = n(), dist = mean(distance), delay = mean(arr_delay)) %>% 
  filter(dest != "HNL", dest != "ANC", count >= 10)
```

### **Exercise #4**

Which carrier had the worst delays over the year? 

First filter out all canceled flights; then group by carrier; next calculate average arrival delay per carrier; last, rank your results. 

Try to use the pipe `%>%` to answer this question.

**ANSWER**

```{r}

flights %>% # first start with flights
  filter(!is.na(dep_delay), !is.na(arr_delay)) %>% # then filter out cancelled flights
  group_by(carrier) %>% # then group by carrier
  mutate(delay = mean(arr_delay)) %>% # then create new column for delay
  arrange(desc(delay)) # and lastly arrange by delay

# Answer: The carrier with the worst delays was F9

```
