---
title: "STAT167 Lab #3 - Spring 2021"
author: 
- name: Deisy Diaz
  email: ddiaz022@ucr.edu
date: "2021/4/16"
output:
  html_document:
    toc: yes
---

# **Discussion week 3 instructions**

This week, we will review some `ggplot2` example figures from the lectures. 

- First, download the `rmd` file from iLearn. 
- Open this `rmd` file in RStudio and click `Knit -> Knit to HTML` to render it to HTML format. (The computer lab computers do not have `LaTex` installed so you can not render it to PDF format).
- Read this `rmd` file and the rendered `html` file side-by-side, to see how this document was generated!
- Be sure to play with this document! Change it. Break it. Fix it. The best way to learn R Markdown (or really almost anything) is to try, fail, then find out what you did wrong.

- Read over the `ggplot2` example code and check the output. If you have any questions about certain functions or parameters, it is the time to ask!
- There are some exercises through out this document. Replace **INSERT_YOUR_ANSWER** with your own answers. Knit the file to HTML, and check your results.

<font color="blue">**Please comment your R code thoroughly, and follow the R coding style guideline (https://google.github.io/styleguide/Rguide.xml). Partial credit will be deducted for insufficient commenting or poor coding styles.**</font>

**Lab submission guideline**

- After you completed all exercises, save your file to `FirstnameLastname-SID-lab3.rmd` and save the rendered html/pdf file to `FirstnameLastname-SID-lab3.html` or `FirstnameLastname-SID-lab3.pdf`.
- Submit **BOTH your source `rmd` file and the knitted `html` or `pdf` file** to **GradeScope**. Do NOT create a zip file. You can submit multiple times, you last submission will be graded.

***

# **Lecture Review - `ggplot2`**

## Install the `tidyverse` package

```{r, collapse=T}
# install the tidyverse package first if you have not done it  yet.
#install.packages("tidyverse") # you can comment out this line after you have installed `tidyverse`
library(tidyverse)
```

## The `mpg` data set

This data set contains fuel economy data 1999 - 2008 for 38 popular car models (source: EPA. [fueleconomy.com/](http://fueleconomy.gov)).

```{r, collapse=T}
?mpg
dim(mpg)  # dimension of the table
mpg  # print/view mpg (we will introduce tibble later in this class)
str(mpg)  # list the structures in mpg
glimpse(mpg) # get a glimpse of the mpg data
```


## The complete graphing template in `ggplot2`
```{r, eval=F}
ggplot(data = <DATA>) + 
  <GEOM_FUNCTION>(
     mapping = aes(<MAPPINGS>),
     stat = <STAT>,        # optional
     position = <POSITION> # optional
  ) +
  <COORDINATE_FUNCTION> +  # optional
  <FACET_FUNCTION> +       # optional
  <SCALE_FUNCTION> +       # optional
  <THEME_FUNCTION>         # optional
```

## Example: aesthetic mappings for both `geom_point()` and `geom_smooth()`

```{r, collapse=T}
ggplot(data = mpg, 
       mapping = aes(x = displ, y = hwy, col = drv) # global mapping
       ) +
  geom_point() +
  geom_smooth(mapping = aes(linetype = drv)) # local mapping for the layer
```

### **Exercise #1**
Write your own `ggplot2` code to make a scatterplot of `cty` against `displ`; add a regression line to your plot; map `drv` to the color aesthetic for both points and lines; map `cyl` to the `size` aesthetic for points. 

**ANSWER:**
```{r, collapse=T}
ggplot(data = mpg, 
       mapping = aes(x = displ, y = cty, col = drv) # global mapping
       ) +
  geom_point(mapping = aes(size = cyl)) +
  geom_smooth(mapping = aes(color = drv)) # local mapping for the layer
```


What you can conclude from your figure?

**ANSWER:** As the cylinder size grows so does displacement but city mileage decreases.


## Facets - making subplots that each display one subset of the data.

### Facets by a single variable - `facet_wrap()`

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ class, nrow = 2)
```

### Facets by a combination of two variables - `facet_grid()`

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(drv ~ cyl)
```

### **Exercise #2**

(a) Look at the following code and output figure. What happens if you facet on a continuous variable?

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = cyl, y = hwy)) + 
  facet_wrap(~ displ, nrow=4)
```

**ANSWER:** If you facet along a continous variable, it will print out points for every single value.

(b) Look at the following code and output figures. What does `.` do?

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(. ~ class)
```

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(class ~ .)
```

**ANSWER:** The dot determines whether the plots are horizontal(when dot is on the right side) or vertical(when dot is on the left side).

## The `diamonds` data set

This data set contains the prices and other attributes of almost 54,000 diamonds. 

```{r, collapse=T}
?diamonds
dim(diamonds)  # dimension of the table
diamonds # print/view diamonds
str(diamonds)  # list the structures in diamonds
glimpse(diamonds) # get a glimpse of the data
```

### **Exercise #3** 
Which variables in `diamonds` are categorical? Which variables are numerical? 

**ANSWER: ** Variables cut, color, clarity, are categorical. The rest are numerical, carat, depth, table, price, x, y, z. 

## geom vs stat functions

Statistical transformation is a alternative way to build a layer.

Most geoms and stats come in pairs that are almost always used in concert. 

- every geom has a default stat
- every stat has a default geom. 

You can typically use geoms without worrying about the underlying statistical transformation. 

You can also use geoms and stats interchangeably.

## Example: `geom_bar()` calls `stat_count()` make bar plot for a categorical variable 

```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut))
ggplot(data = diamonds) + 
  stat_count(mapping = aes(x = cut))
```

## Override default stat function

To make a proportion bar chart (relative frequency histogram), we need to override the default `group` argument and map the variable `..prop..` (calculated by `stat_count() `) to the `y` aesthetic.

```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, y = ..prop.., group = 1))
```


## Example: position adjustment options for `geom_bar()` 

```{r, collapse=T}
?geom_bar
```

The position argument specifies the position adjustment of bars, rectangles.

- **default**: position = "stack"
- position = "identity" will place each object exactly where it falls in the context of the graph. 
- position = "fill" works like stacking, but makes each set of stacked bars the same height. 
- position = "dodge" places overlapping objects directly beside one another. 
the bars are automatically stacked. Each colored rectangle represents a combination of cut and clarity.


```{r, collapse=T, fig.height=5}
#install.packages("gridExtra") # you only need to install it once
library(gridExtra)
gg <- ggplot(data = diamonds,
             mapping = aes(x = cut, fill = clarity)) 
plot1 <- gg + geom_bar(position="stack")
plot2 <- gg + geom_bar(position="identity")
plot3 <- gg + geom_bar(position="fill")
plot4 <- gg + geom_bar(position="dodge")
grid.arrange(plot1, plot2, plot3, plot4, ncol=2)
```
